BOLD Parameter Estimation Using Particle Filters

### 2010-08-07 doc/thesis/background.tex
fixes in methods and results. Also big additions to conclusion

\subsection{Linear Approximation}
In \cite{Friston2002b}, a novel combination of linear and nonlinear modeling
was used to generate parameter estimates. Because there is no closed 
form solution to the ballon model, it impossible to 
calculate the Jacobian matrix $\frac{\partial Y}{\theta}$. Thus 
\cite{Friston2002b} approximates the partial using a Volterra Kernel. 
At each step of the Expectation-
Maximization (EM) algorithm, the differential equation is integrated, calculating
the residuals. Then, for each $\theta_i$ surrounding the current estimate of 
$\theta$, a Volterra-Kernel expansion of the output $y$ is generated. Generation
of the Volterra Kernel is quick, and, since there
is an analytical solution calculating partial derivatives is easy. 
 The full E-M algorithm
for estimating the states is notation-heavy and can be found in \cite{Friston2002b}.

@@

\subsection{Nonlinear Least Squares}
\label{sec:Nonlinear Least Squares}
Although there are certainly benefits to using a derived model, rather
than a purely empirical model, there are serious implications. The
first problem is that all the powerful classical techniques of gradient
descent are off limits; since the model is a true nonlinear dynamical
system with no closed form solution (although \autoref{sec:Linear Approximation}
circumvented this by calculating a Volterra Series approximation). Without
a Jacobian for residuals, the Gauss-Newton method
is impossible. Additionally, a gradient descent is slow
without the ability to form partials of the output with respect
to all the parameters.

@@

>>>>

\subsection{Polynomial Approximation}
\label{sec:Background Linear Approximation}
In \cite{Friston2002b}, a novel combination of linear and nonlinear modeling
was used to generate parameter estimates. Because there is no closed 
form solution to the ballon model, it impossible to 
calculate the Jacobian matrix $\frac{\partial Y}{\theta}$. Thus 
\cite{Friston2002b} approximates the partial using a Volterra Kernel. 
At each step of the Expectation-
Maximization (EM) algorithm, the differential equation is integrated, calculating
the residuals. Then, for each $\theta_i$ surrounding the current estimate of 
$\theta$, a Volterra-Kernel expansion of the output $y$ is generated. Generation
of the Volterra Kernel is quick, and, since there
is an analytical solution calculating partial derivatives is easy. 
 The full E-M algorithm
for estimating the states is notation-heavy and can be found in \cite{Friston2002b}.

@@

\subsection{Nonlinear Least Squares}
\label{sec:Nonlinear Least Squares}
Although there are certainly benefits to using a derived model, rather
than a purely empirical model, there are serious implications. The
first problem is that all the powerful classical techniques of gradient
descent are off limits; since the model is a true nonlinear dynamical
system with no closed form solution (although \autoref{sec:Background Linear Approximation}
circumvented this by calculating a Volterra Series approximation). Without
a Jacobian for residuals, the Gauss-Newton method
is impossible. Additionally, a gradient descent is slow
without the ability to form partials of the output with respect
to all the parameters.

@@

\subsection{Previous Particle Filter Approaches}
In \cite{Murray2008} a particle filter based approach was used to integrate
the BOLD equations. The method used in that work focused primarily on estimating
the BOLD output and state equations as a nonlinear stochastic differential 
equation. The primary difference between \cite{Murray2008} and this work is that
\cite{Murray2008} took the parameters as a given. Thus, differences in the BOLD output
were taken to be primarily driven by stochastic changes in the underlying state
equations. Because the parameters were not allowed to change, the estimate of 
the BOLD signal was not very good. The fact that the
differences in BOLD response cannot be explained solely by stochastic elements 
is important, however. The particle filter library created
in that work, $dysii$, forms the basis for the particle filter used in this work, 
and was extremely well designed. The work also clearly presents the particle filter;
both its derivation and use.

### 2010-08-07 doc/thesis/conclusion.tex
fixes in methods and results. Also big additions to conclusion

* Overview of results
** Parameters under-constrained
** Output estimates good

* Pros/Cons
** Pro
*** Non-parametric
*** Real time
*** Full Posterior
*** More intuitive
** Cons
*** Computation longer than SPM
*** Harder to interpret

* Enhancements/Dehancements 
** flatten prior
** \% difference using spline?
** "smart knots"
** DC gain as a parameter
** linearizing

* Future works
** Extensive test of quality of volterra kernel estimate from friston
** Better de-trending
** Automatic estimation of measurement error.

@@

Automatic detection of the noise level in the signal, to get a decent wieghting
function. Large scale activation to get more parameter estimates. 
Viscoelastic effects from \cite{Buston2004}, discussed in \cite{Deneux2006}.

>>>>

% Overview of results
%% Parameters under-constrained
%% Output estimates good
\section{Review of Results}
The results unequivocally show that the parameters of the BOLD model are under
constrained. While unsurprising given the sensitivity analysis in \cite{Deneux2006},
it is still an important limitation when calculating parameters. Thus, other methods
that depend on a single point estimate of the parameters, such as kalman filters
or even least squares are limited to estimating the BOLD signal, but the estimate
of the underlying parameters and state variables are suspect at best. Similarly, as the
histograms in \autoref{sec:Multi-voxel Simulation} and \autoref{sec:Real Data Parameter Estimates}
show, using the mean to estimate parameters also makes little sense in the particle filter
case. Attributing the variance of estimate to the underlying parameter distribution would
also be a mistake. While it is definitely true that parameters vary from person to person
and region to region, the estimated distributions in \autoref{sec:Real Data Parameter Estimates} 
are smoothed versions of those distributions. For this reason, any analysis of parameters should
take the entire estimated posterior distribution into account, and perform F-tests on that
distribution, rather than an idealized Gaussian. In this sense, particle filters represent
an important step forward in BOLD parameter estimation. It is now clear that representing
the uncertainty in parameters as a Gaussian is insufficient; so using a particle filter
or Bayesian estimate of the posterior is not simply an enhancement, but a necessary precaution.

Although point estimates of parameters are not dependable estimates of the 
true parameters, they are by no means usefulness. Analysis of differences in parameters
could be medically relevant. Additionally, a single estimate for parameters is still able
to form a good estimate of the BOLD signal. This factor adds bridge to 
earlier types of activation tests, such as the statistical
parametric mapping. As the results in \autoref{sec:RealData} show, the heatmaps, especially
those of mutual information closely resembled the results of SPM. While the activation
tests were more sensitive, they were also more prone to false positive. However, it is
worth noting that certain enhancements could be useful in reducing false positives; for instance
by using the maximum likelihood of the final distribution, or even the median. The 
usefulness of these techniques depend on the degree to which the algorithm converged, 
and thus on the experimental design. Future work may shed further light on these
techniques.

% Pros/Cons
%% Cons
%%% Computation longer than SPM
%%% Harder to interpret
%% Pro
%%% Non-parametric
%%% Real time
%%% Full Posterior
%%% More intuitive
\section{Particle Filter Approach}
The Particle Filter algorithm was originally designed for on-line parameter 
estimation. For this reason, there is no guarantee of optimality or even 
convergence for finite measurements. However, for the BOLD nonlinear ODE there
can be no guarantee of an optimal solution. A guaranteed local minimum,
which nonlinear least squares provides, however would be helpful. One difficulty
with the use of a particle filter with finite data is the calculation of a good
weight function $P(y_k | x_k)$  that will converge at a decent rate. If the weight
function does not sufficiently differentiate particles, the final distribution will 
no be significantly different from the prior distribution. On the other hand, if the 
weight function is too thin, it will unfairly eliminate viable particles. Given sufficient
measurements it is better to let the algorithm take longer to converge, because the
convergence will be better (more accurate). The particle filter takes longer
to run than Volterra approximation method from \autoref{sec:Background Linear Approximation};
however, it is free from the uncertainty of whether a quadratic approximation is 
sufficient for the BOLD model.

The particle filter certainly has significant advantages over other estimation procedures
discussed in \autoref{sec:Prior Works}. The most important advantage is that it provides
an estimate of the posterior probability, rather than a single estimate. While it is natural
to want a simple estimate of parameters, such an estimate is impossible with this particular
model. The results are more difficult to interpret, but this is a necessity. The fact that 
the final distribution is not dependent on any particular distribution is also advantageous.
Of particular note; the final distribution does not need to conform to any parametric
distribution. While the particle filter was not fast for full brain calculations, its speed
was sufficient on a quad core machine to perform real time calculations of small regions
(approximate run time .27 seconds per voxel-measurement). Today it would be possible
 to perform real time analysis of 10 voxels on an average quad core. The algorithm also scales
well and does not require burdensome amounts of memory (approximately 11 megabytes). 
For this reason this algorithm is perfect for extension to vector or video card
based processors.

A more practical concern with the particle filter is that it is mathematically
simple. Only a basic understanding of Bayesian statistics is needed to understand
how the particle filter works. This is in contrast to the Volterra based approach
of \cite{Friston2000}, which is quite complex. Additionally very few assumptions
are needed for the particle filter. It is a common problem in traditional statistics 
for assumptions to be unrealistic which means the results may be invalid or at 
least in question. 
Fewer and more realistic assumptions mean that the particle filter is more robust
to unforeseen difficulties in FMRI data.

% Enhancements/Dehancements 
%% flatten prior
%% \% difference using spline?
%% "smart knots"
%% DC gain as a parameter
%% linearizing

% Future works
%% Extensive test of quality of volterra kernel estimate from friston
%% Better de-trending
%% Automatic estimation of measurement error.

@@

Automatic detection of the noise level in the signal, to get a decent wieghting
function. Large scale activation to get more parameter estimates. 
Viscoelastic effects from \cite{Buxton2004}, discussed in \cite{Deneux2006}.

### 2010-08-07 doc/thesis/methods.tex
fixes in methods and results. Also big additions to conclusion

\section{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied regions deemed active by the General 
Linear Model; and most other research (including \cite{Friston2002b}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to a extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. A very wide prior will require
more particles to be sufficiently dense, a very thin prior may miss
the true parameters. Consequently, for this work it was natural
to use priors that will give results consistent with similar tests. accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

>>>>

\section{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied regions deemed active by the General 
Linear Model; and most other research (including \cite{Friston2001}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to a extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. A very wide prior will require
more particles to be sufficiently dense, a very thin prior may miss
the true parameters. Consequently, for this work it was natural
to use priors that will give results consistent with similar tests. accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

### 2010-08-07 doc/thesis/results.tex
fixes in methods and results. Also big additions to conclusion

\section{Multi-voxel Simulation}
To test the usefulness of the particle filter for I used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As an unfortunate side effect of 
not using Possums' original activation scale, I manually added 750 to the total level of
simulated Possum images. This is because the BOLD \% difference levels were in the range
of 50 - 100\% from the base, about 5 times as large as they should have been. Ultimately
this has no effect on the quality of the simulation and should not effect the 
underlying parameters (other than perhaps $\epsilon$ and $V_0$). 
Also worth noting is that the noise level was set to an SNR of 20, due to changes in the 
program the true signal-to-noise ratio was much lower, as seen in the fit in \autoref{fig:simslicefit1}.

>>>>

\section{Multi-voxel Simulation}
\label{sec:Multi-voxel Simulation}
To test the usefulness of the particle filter for I used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As an unfortunate side effect of 
not using Possums' original activation scale, I manually added 750 to the total level of
simulated Possum images. This is because the BOLD \% difference levels were in the range
of 50 - 100\% from the base, about 5 times as large as they should have been. Ultimately
this has no effect on the quality of the simulation and should not effect the 
underlying parameters (other than perhaps $\epsilon$ and $V_0$). 
Also worth noting is that the noise level was set to an SNR of 20, due to changes in the 
program the true signal-to-noise ratio was much lower, as seen in the fit in \autoref{fig:simslicefit1}.

### 2010-08-07 doc/thesis/resultsB.tex
fixes in methods and results. Also big additions to conclusion

\section{Parameter Estimates}
Although the parameters are not uniquely identifiable by a single time-series, that
does not mean estimating them is not without benefit. The parameters still contain useful
information about the system. Additionally, as an aggregate they form a distribution of
the feasible parameters parameters for a particular patient. Therefore 
\autoref{fig:pmap0} through \autoref{fig:pmap6}
contain the parameter maps for the system and \autoref{fig:realhist} is a histogram across all voxels
for which the mutual information was greater than $.15$. As before, the threshold is not
scientifically derived, yet in tests this threshold provided a decent balance to remove most
of the questionably active voxels in the system.

>>>>

\section{Parameter Estimates}
\label{sec:Real Data Parameter Estimates}
Although the parameters are not uniquely identifiable by a single time-series, that
does not mean estimating them is not without benefit. The parameters still contain useful
information about the system. Additionally, as an aggregate they form a distribution of
the feasible parameters parameters for a particular patient. Therefore 
\autoref{fig:pmap0} through \autoref{fig:pmap6}
contain the parameter maps for the system and \autoref{fig:realhist} is a histogram across all voxels
for which the mutual information was greater than $.15$. As before, the threshold is not
scientifically derived, yet in tests this threshold provided a decent balance to remove most
of the questionably active voxels in the system.
